\textsc{Beyond the Iceberg Hypothesis: \\Opening the Black Box of Transport Costs}

### 2017-05-17 redaction/dhp_sept2016.tex
Changements redaction

\date{September 2016}
 \maketitle
\bigskip

@@

Defined as the costs associated with the exchange of goods across national borders, trade costs are usually split into transaction costs (information costs, contract enforcement costs, costs associated with the use of different currencies...), policy costs (tariff  and non-tariff costs), time costs (time to ship goods) and transport costs \emph{per se}. In this vein, \citet{anderson_wincoop_jel} obtain that the 74\% markup made by international trade costs can be divided between 44\% for border-related trade barriers and 21\% for transportation costs. This last figure suggests that the share of transport costs in the consumer price of manufactured goods is substantial.
\citet{Behar_Venables} obtain that the elasticity of trade with respect to freight costs is sizeable, by around -3. If much of trade-policy barriers have been removed over the second half of the 20th century, these findings point out that the transport costs component of the overall trade costs remain large and deserve attention. This is accordingly the focus of the paper.\smallskip

\todo{reprendre slides presentation LEDa pour la motivation de la question.}
\todo{JH: apparently, it has been done! Or do you want to have the Italian shoes example in the intro????}

Following \citet{samuelson1954}, standard models of international trade have usually relied on modelling trade costs as an \emph{ad-valorem} tax equivalent (ie, as a constant percentage of the producer price per unit traded, part of the ``iceberg cost'' hypothesis\footnote{Rigourously speaking, the ``iceberg cost'' includes the ad-valorem costs and the fact that the costs are paid in terms of the good that is traded. As this last element is irrelevant in our case, we use the terme ``iceberg'' and ``ad-valorem'' interchangeably, as commonly made in the literature.}). However, many common empirical facts support the existence of additive costs. As documented by \citet{Irrazabal_2015}, pricing structure in shipping, additive tariffs, distribution costs... often exhibit (at least partly) an additive structure. The structure (additive vs ad-valorem) of transport costs is far from being anecdotal, as the literature has long emphasized its role in shaping the pattern of trade flows. The Alchian and Allen conjecture (\citealp{alchian}), that points out that the relative price of two varieties of some good will depend on the level of trade costs, does rely on the existence of additive costs: The relative demand for more expensive/higher quality product goods should increase with trade cost (``shipping the good apples out''). \citet{Lashkaripour-mimeo-2016} challenges this view. He finds supporting evidence for the \emph{ad-valorem} assumption by taking into account the fact that more expensive goods are systematically heavier and hence more costly to transport. One can yet be concerned by the generality of this result. By nature, his study is restricted to goods that are enumerated by items in the statistics (they represent 60\% of US imports). Furthermore, while the positive correlation between weight and price seems reasonable for goods from the second industrial revolution like cars, it is dubious in the case of ITC goods which importance has been rising since 1994 (the end point of Lashkaripour's study). Besides, a number of empirical papers provide a strong empirical support to the role of additive costs in international costs. Based on a firm-product-level database of French exporters, \citet{martin2012} finds that firms charge higher fas unit values on exports to more remote countries, in contradiction with the ad-valorem hypothesis. \citet{hummels_skiba} estimate the elasticity of freight rates with respect to price to be well below unity. Also, their estimates imply that doubling freight costs increases average fas export prices by 80-141 percent, consistent with high quality goods being sold in markets with high freight costs. These findings deliver strong empirical support in favor of the Alchian-Allen conjecture.
\smallskip

@@

@@

Third, we exploit the time dimension of our database to document the patterns of transport costs over time. As pointed out by \cite{hummels2007}, we confirm the importance of the composition effects by country of origin and by product in shaping the patterns of transportation costs over time. After excluding the composition effects, we find that all types of transport costs have been roughly constant from 1974 to 1984 and then steadily decreased by 40\% over the period 1984-2013. Further, our results point out an important bias in the time trends of transport costs when additive costs are omitted. While transport costs seem to exhibit a larger decrease in air than in sea transport between 1985 and 2005 when only iceberg costs are modeled, this difference vanishes when additive costs are taken into account. These results therefore confirm the importance of the additive component in accounting for international transport costs.\smallskip

@@

@@

\todo{Comparer avec la litterature. Cf AvW 2004 pure transport costs 11\% markup over production costs.}

@@

Before making any definite statement about this, it is worth emphasizing that the evolution of transport costs depends on the one hand of the evolution of per product and per partner costs and on the other hand of the evolution of the composition of trade.
Total transport costs may have decreased over time because the share of neighbour countries in total US trade or the share of cheaper to transport goods have increased, independently of any change in transport costs \textit{per se}. As argued by \citet{hummels2007}, it is hence necessary to eliminate the composition effects of trade flows to isolate the evolution of per product- and per partner- transport costs.

\todo{a dire pas forcement ici, tout n'est pas que sur cette question des effets de composition.}

\todo{argument general}
In comparison with \citet{hummels2007}, our empirical analysis allows for a more flexible estimation of the share between the additive and the ad-valorem components of transport costs. Precisely, we allow it to change over time and across country partners and products, while \citet{hummels2007} imposes it to be constant in these three dimensions.\footnote{In the terms of \cite{hummels_skiba}, our estimation strategy allows for a varying $\beta$ across time, country partner and product.}

\todo{argument sur les effets de composition}
If we share this view with \citet{hummels2007}, our strategy for eliminating composition effects differs from his in two main respects. First, we provide a broader characterization of the time trends in transport costs as our empirical analysis allows for a more flexible estimation of the share between the additive and the ad-valorem components of transport costs. In characterizing the time trend in the ``pure transport costs'', \citet{hummels2007} eliminates the changes in the price per kilo dimension, attributing this to a change in the composition of the bundle over time (per country-commodity).\footnote{See \citet{hummels2007}, p. 146.} By contrast, we provide an explicit decomposition of this additive cost component over time: By comparing the additive cost component unfitted (total) and fitted (composition effects excluded), we are thus able to characterize if this dimension of the transport cost has fallen due to (for instance) the fact that product quality has increased over time (weight has reduced, hence reducing the transport cost by dollar transported), or if it is the ``pure'' additive costs (for instance, handling costs) that have reduced over time. As well, we are able to do so for the multiplicative component of trade costs. We also agglomerate the two components in a unified measure of transport costs from which we exclude the composition effects. This allows us to compare how ``pure'' transport costs have changed over time, composition effects excluded, and to compare them to the unfitted ones, reported in Figure \ref{fig:decomp_TC_3d}). Last, for sake of comparison though, we do the same exercise as \citet{hummels2007}.

>>>>

\date{May 2017}
 \maketitle
\bigskip

@@

Defined as the costs associated with the exchange of goods across national borders, trade costs are usually split into transaction costs (information costs, contract enforcement costs, costs associated with the use of different currencies...), policy costs (tariff  and non-tariff costs), time costs (time to ship goods) and transport costs \emph{per se}. In this vein, \citet{anderson_wincoop_jel} obtain that around 30\% of international trade costs are attributable to transport costs. Equivalently, international transportation costs would represent a 21\% markup over production costs. \citet{Behar_Venables} obtain that the elasticity of trade with respect to freight costs is sizeable, by around -3, suggesting a non-negligible impact of transport costs on trade flows. If much of trade-policy barriers have been removed over the second half of the 20th century, these findings point out that the transport costs component of the overall trade costs remain large and deserve attention. This is accordingly the focus of the paper.\smallskip

Following \citet{samuelson1954}, standard models of international trade have usually relied on modelling trade costs as an \emph{ad-valorem} tax equivalent (ie, as a constant percentage of the producer price per unit traded, part of the ``iceberg cost'' hypothesis\footnote{Rigourously speaking, the ``iceberg cost'' includes the ad-valorem costs and the fact that the costs are paid in terms of the good that is traded. As this last element is irrelevant in our case, we use the terms ``iceberg'' and ``ad-valorem'' interchangeably, as commonly made in the literature.}). However, many common empirical facts support the existence of additive costs. As documented by \citet{Irrazabal_2015}, pricing structure in shipping, additive tariffs, distribution costs... often exhibit (at least partly) an additive structure. The structure (additive vs ad-valorem) of transport costs is far from being anecdotal, as the literature has long emphasized its role in shaping the pattern of trade flows. The Alchian and Allen conjecture (\citealp{alchian}), that points out that the relative price of two varieties of some good will depend on the level of trade costs, does rely on the existence of additive costs: The relative demand for more expensive/higher quality product goods should increase with trade cost (``shipping the good apples out''). \citet{Lashkaripour-mimeo-2016} challenges this view. He finds supporting evidence for the \emph{ad-valorem} assumption by taking into account the fact that more expensive goods are systematically heavier and hence more costly to transport. One can yet be concerned by the generality of this result. By nature, his study is restricted to goods that are enumerated by items in the statistics (they represent 60\% of US imports). Furthermore, while the positive correlation between weight and price seems reasonable for goods from the second industrial revolution like cars, it is dubious in the case of ITC goods which importance has been rising since 1994 (the end point of Lashkaripour's study). Besides, a number of empirical papers provide a strong empirical support to the role of additive costs in international costs. Based on a firm-product-level database of French exporters, \citet{martin2012} finds that firms charge higher fas unit values on exports to more remote countries, in contradiction with the ad-valorem hypothesis. \citet{hummels_skiba} estimate the elasticity of freight rates with respect to price to be well below unity. Also, their estimates imply that doubling freight costs increases average fas export prices by 80-141 percent, consistent with high quality goods being sold in markets with high freight costs. These findings deliver strong empirical support in favor of the Alchian-Allen conjecture.
\smallskip

@@

\todo{mettre une phrase pour dire qu'on confirme la litterature sur le role des additifs, mais de maniere differente. Hummels et Skiba sur des donnes de 1994, on va beaucoup plus loin qu'eux dans cette voie.}

@@

Third, we exploit the time dimension of our database to document the patterns of transport costs over time. As pointed out by \cite{hummels2007}, we confirm the importance of the composition effects by country of origin and by product in shaping the patterns of transportation costs over time. In contrast to his work though, our empirical analysis allows for a more flexible estimation of the share between the additive and the ad-valorem components of transport costs. Precisely, we allow it to change over time and across country partners and products, while \citet{hummels2007} imposes it to be constant in these three dimensions. After excluding the composition effects, we find that all types of transport costs have been roughly constant from 1974 to 1984 and then steadily decreased by 40\% over the period 1984-2013. Further, our results point out an important bias in the time trends of transport costs when additive costs are omitted. While transport costs seem to exhibit a larger decrease in air than in sea transport between 1985 and 2005 when only iceberg costs are modeled, this difference vanishes when additive costs are taken into account. These results therefore confirm the importance of the additive component in accounting for international transport costs.\smallskip

@@

\todo{expliciter ici notre strategie d'estimation par rapport a Hummels et Skiba, 2004. In the terms of \cite{hummels_skiba}, our estimation strategy allows for a varying $\beta$ across time, country partner and product.}

@@

@@

Before making any definite statement about this, it is worth emphasizing that the time trend of international transport costs depends on both \textit{i°)} the evolution of per product and per partner transport costs and \textit{ii°)} the evolution of the composition of trade. Total transport costs may thus have decreased over time because the share of neighbour countries in total US trade or the share of goods cheaper to transport has increased (explanation \textit{ii°)}), independently of any change in transport costs \textit{per se}. As argued by \citet{hummels2007}, it is hence necessary to eliminate the composition effects of trade flows to isolate the evolution of ``pure'' international transport costs, i.e. per product- and per partner- transport costs.

If we share this view with \citet{hummels2007}, our strategy for eliminating composition effects differs from his in two main respects. First, we provide a broader characterization of the time trends in transport costs as our empirical analysis allows for a more flexible estimation of the share between the additive and the ad-valorem components of transport costs. In characterizing the time trend in the ``pure transport costs'', \citet{hummels2007} eliminates the changes in the price per kilo dimension, attributing this to a change in the composition of the bundle over time (per country-commodity).\footnote{See \citet{hummels2007}, p. 146.} By contrast, our estimation strategy allows for an explicit decomposition of the additive cost component over time/product/country partner. \textbf{stop here} By comparing the additive cost component unfitted (total) and fitted (composition effects excluded), we are thus able to characterize if this dimension of the transport cost has fallen due to (for instance) the fact that product quality has increased over time (weight has reduced, hence reducing the transport cost by dollar transported), or if it is the ``pure'' additive costs (for instance, handling costs) that have reduced over time. As well, we are able to do so for the multiplicative component of trade costs. We also agglomerate the two components in an unified measure of transport costs from which we exclude the composition effects. This allows us to compare how ``pure'' transport costs have changed over time, composition effects excluded, and to compare them to the unfitted ones, reported in Figure \ref{fig:decomp_TC_3d}). Last, for sake of comparison though, we do the same exercise as \citet{hummels2007}.
